<!-- question-type: inclass -->

### Exercise 7: Producing Evidence that Answers Your Question

Using the dataset your group prepared in Exercise 6, produce **one or two graphs or tables** that help answer your allocated question.

Your output should match the scope your group chose in Exercise 5.

::: {.callout-note collapse="true"}

### What should a good output make clear?

Your graph or table should be clear enough that another group could understand:

* what variables you are showing
* what subset of the data you chose to use
* what pattern or relationship the output is meant to highlight

:::

::: {.callout-note collapse="true"}

### Presentation reminders

If you create a graph, make sure it has informative axis labels and a clear title.
If you create a table, make sure it has clear column names and is easy to interpret.
:::

::: {.callout-note collapse="true"}
### Hint: plotting functions you may find useful

You will not need all of these, but they may help get you started: `ggplot()`, `aes()`, `geom_point()`, `geom_histogram()`, `geom_bar()`, `geom_line()`, `geom_smooth()`, `facet_wrap()`, `scale_x_log10()`, `labs()`, `theme_minimal()`.

:::

<!-- BEGIN PROFILE:r-teaching-guide -->

::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

⏱️ **Timing**

~15 minutes. Groups should complete AT LEAST ONE presentation-ready plot/table. Ideally, two. 

🎯 **Learning Objective**

Students should:

* turn a cleaned analysis dataset into evidence that speaks to their allocated question
* choose an output format that fits the question and scope they selected earlier
* present a graph or table clearly enough for another group to interpret
* connect data wrangling decisions to communication choices

✅ **Core Concepts to Highlight**

* A good output should help answer the question, not just display the data.
* Students should choose the format that best communicates the relationship they want to show.
* A graph or table becomes more useful when the title, labels, and scope are clear.
* Simpler outputs are often better than overloaded ones.
* This exercise is about producing evidence that can later be used in the short brief.

💬 **Suggested In-Class Prompts**

* What does this graph or table help us see?
* Why did you choose this format rather than another one?
* Does the output match the scope you chose in Exercise 5?
* Could another group understand what you found without extra explanation?
* Are there any labels, titles, or notes missing?

📌 **Common Misunderstandings**

* Students may produce a graph or table that is visually busy but does not answer the question clearly.
* Students may forget to match the output to the countries, years, or variation they selected earlier.
* Students may leave axes, titles, or column names too vague.
* Students may assume graphs are always better than tables.
* Students may create two outputs that duplicate each other rather than adding insight.

:::

:::

:::

<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->

::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

There are many possible good answers to this exercise. What matters most is that the graph or table helps answer the group's question clearly.

Below is **one possible solution** that continues the example from Exercise 6 for the question **How does urbanisation relate to GDP per capita?**

One sensible approach is to create:

1. a graph showing the relationship between GDP per capita and urbanisation for selected years
2. a table summarising continent-level averages for a few benchmark years

In this case, we would usually **prefer the graph** because the main question is about the **relationship** between two variables, and a graph makes that pattern easier to see. That said, if a group is pressed for time, a clear summary table can still be a valid alternative.

```{r}
#| eval: false

# Graph: relationship between GDP per capita and urbanisation
analysis_df |>
  filter(year %in% c(1960, 1990, 2020)) |>
  ggplot(
    aes(
      x = gdp_pcap_pp / 1e3,
      y = urban_pct,
      size = pop,
      colour = continent_name
    )
  ) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    aes(x = gdp_pcap_pp / 1e3, y = urban_pct),
    method = "lm",
    formula = y ~ poly(x, 2),
    linetype = "dashed",
    se = FALSE,
    colour = "grey40"
  ) +
  scale_x_log10() +
  ggokabeito::scale_color_okabe_ito() +
  scale_size(guide = "none") +
  facet_wrap(~year) +
  labs(
    title = "Urbanisation and GDP per capita across countries",
    subtitle = "Selected years: 1960, 1990, and 2020",
    x = "GDP per capita ('000s of USD, 2021 PPP adjusted)",
    y = "Population living in urban areas (%)",
    colour = "Continent"
  ) +
  theme_bw()
```

```{r}
#| eval: false
# Table: continent-level summary for the same benchmark years
table_continent_year <- analysis_df |>
  filter(year %in% c(1960, 1990, 2020)) |>
  group_by(continent_name, year) |>
  summarise(
    avg_gdp_pcap_pp = weighted.mean(gdp_pcap_pp, w = pop, na.rm = TRUE),
    avg_urban_pct   = weighted.mean(urban_pct, w = pop, na.rm = TRUE),
    total_pop       = sum(pop, na.rm = TRUE),
    n_countries     = n_distinct(geo),
    .groups = "drop"
  ) |>
  mutate(
    avg_gdp_pcap_pp = round(avg_gdp_pcap_pp, 0),
    avg_urban_pct   = round(avg_urban_pct, 1),
    total_pop_m     = round(total_pop / 1e6, 1)
  ) |>
  select(
    continent = continent_name,
    year,
    n_countries,
    total_pop_m,
    avg_gdp_pcap_pp,
    avg_urban_pct
  ) |>
  arrange(continent, year)

table_continent_year |>
  kable(
    col.names = c(
      "Continent",
      "Year",
      "Countries",
      "Population (millions)",
      "Avg GDP per capita",
      "Avg urbanisation (%)"
    ),
    caption = "Urbanisation and GDP per capita by continent, 1960, 1990, and 2020"
  )
```

This is only one possible solution. A group could also choose a simpler graph, a simpler table, a different subset of years, or a different way of comparing countries or regions.

What makes this a strong answer is that:

* the outputs match the question being asked
* the scope is clear and consistent across the graph and table
* the titles and labels help another group interpret the results
* the graph is well suited to showing the relationship between two variables
* the table is a useful fallback or complement when time is short or when a summary by continent is easier to interpret

:::

:::

:::

<!-- END PROFILE:r-solutions -->
