<!-- question-type: inclass -->

### Exercise 6: Build Your Analysis Dataset

Work with your group to build the dataset you need to answer your allocated question. You should decide for yourselves how to prepare the data for analysis.

::: {.callout-note collapse="true"}
### Hint

Depending on your question, you may need to think about some of the following:

* which files you need to load
* whether any datasets need tidying before they can be analysed
* whether any datasets need to be joined together
* whether you want to focus on particular countries, regions, or years
* whether there are missing values that affect the years or observations you can analyse
* what one row in your final analysis dataset should represent

:::

::: {.callout-tip collapse="true"}

### Hint: Reshaping year columns

Some of the datasets store the **year inside the column names** rather than in a separate column. In cases like this, `pivot_longer()` can help you move from a wide format to a tidy format.

For example, if your GDP dataset has columns such as `gdp_pcap_pp_1960`, `gdp_pcap_pp_1961`, and so on, you could reshape it like this:

```{r}
#| eval: false
gdp <- gdp |>
  pivot_longer(
    cols = starts_with("gdp_pcap_pp_"),
    names_to = "year",
    names_pattern = "gdp_pcap_pp_(\\d{4})$",
    values_to = "gdp_pcap_pp"
  )
```

Here is what each part is doing:

* `cols = starts_with("gdp_pcap_pp_")` selects the year columns you want to reshape
* `names_to = "year"` creates a new column called `year`
* `names_pattern = "gdp_pcap_pp_(\\d{4})$"` extracts the 4-digit year from the original column names
* `values_to = "gdp_pcap_pp"` puts the GDP values into a single column

After this step, the dataset will be much closer to tidy form, where each row can represent one **country-year** observation.
:::

<!-- BEGIN PROFILE:r-teaching-guide -->

::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective**

Students should:

* build a usable analysis dataset from multiple raw files
* tidy wide datasets into a form where each row is one observation
* join multiple datasets using common keys
* inspect missingness and make sensible decisions about scope
* prepare a clean dataset for later visualisation or tabulation

✅ **Core Concepts to Highlight**

* Students need to decide what one row in the final dataset should represent. Here, the natural goal is one **country-year**.
* The messy files are untidy because **year** is stored in the column names.
* Tidying is not separate from analysis. It is what makes joining and plotting possible.
* Missingness often becomes visible only after datasets are joined.
* A sensible analysis sometimes means restricting attention to a period with better data coverage.
* There is more than one workable path, but the final dataset should match the question the group is trying to answer.

💬 **Suggested In-Class Prompts**

* What should one row represent in your final dataset?
* Can you join these datasets in their current form?
* Which variables do you actually need for your question?
* Where do missing values appear after the join?
* Are there years with especially weak coverage?
* What filtering choice would make the later graph or table clearer?

📌 **Common Misunderstandings**

* Students may try to join the files before tidying them.
* Students may tidy one file but forget that the other wide files need the same treatment.
* Students may ignore missing values and move straight to plotting.
* Students may drop missing values without checking whether this removes large parts of the time period.
* Students may build a dataset with the wrong unit of observation.

:::

:::

:::

<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->

::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

There are different valid ways to build the analysis dataset. What matters most is that the final dataset is tidy, joined correctly, and suitable for the group’s question.

Here is one strong example for the question relating **urbanisation** to **GDP per capita**.

```{r}
#| eval: false
# Load the raw files
gdp <- read_csv("out_data/gdp.csv")
urban <- read_csv("out_data/urban.csv")
pop <- read_csv("out_data/pop.csv")
codes <- read_csv("out_data/country_to_continent.csv")

# Tidy GDP:
# the year is currently stored inside the column names,
# so we reshape to make year a variable
gdp <- gdp |>
  pivot_longer(
    cols = starts_with("gdp_pcap_pp_"),
    names_to = "year",
    names_pattern = "gdp_pcap_pp_(\\d{4})$",
    values_to = "gdp_pcap_pp"
  ) |>
  mutate(year = as.numeric(year))

# Tidy urbanisation:
# again, reshape so each row is one country-year
urban <- urban |>
  pivot_longer(
    cols = starts_with("urban"),
    names_to = "year",
    names_pattern = "urban_pct_(\\d{4})$",
    values_to = "urban_pct"
  ) |>
  mutate(year = as.numeric(year))

# Tidy population:
# this can be useful later for bubble size or weighted summaries
pop <- pop |>
  pivot_longer(
    cols = starts_with("pop"),
    names_to = "year",
    names_pattern = "pop_(\\d{4})$",
    values_to = "pop"
  ) |>
  mutate(year = as.numeric(year))

# Join the tidied files together.
# After this step, each row should represent one country-year.
analysis_df <- gdp |>
  left_join(
    urban |> select(-name),
    by = join_by(geo, year)
  ) |>
  left_join(
    pop |> select(-name),
    by = join_by(geo, year)
  ) |>
  left_join(
    codes |> select(-country_name),
    by = join_by(geo == iso3)
  )

# Check data coverage by year.
# This helps us see whether some years have much more missing data than others.
analysis_df |>
  group_by(year) |>
  summarise(
    n_rows = n(),
    pct_complete = mean(
      !is.na(gdp_pcap_pp) &
      !is.na(urban_pct) &
      !is.na(pop) &
      !is.na(continent_name)
    ) * 100
  )

# Suppose we decide that coverage is much better from 1960 onward.
# We can then restrict the time window and remove incomplete rows.
analysis_df <- analysis_df |>
  filter(between(year, 1960, 2020)) |>
  drop_na()

# Inspect the final analysis dataset
glimpse(analysis_df)
```

This is a strong workflow because it does four things clearly:

1. It reshapes the messy files into tidy form.
2. It joins them into one country-year dataset.
3. It checks missingness before moving to analysis.
4. It makes an explicit filtering decision based on data coverage.

The key missingness lesson here is that dropping missing values should usually come **after** you inspect where the gaps are. A better workflow is:

* build the joined dataset
* check coverage
* decide on a sensible analysis window
* then remove incomplete observations if needed

:::

:::

:::

<!-- END PROFILE:r-solutions -->
